Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2021-09-26 Manuscript/SS_GAM.tex

\abstract{There are proposals that extend the classical generalized
additive models (GAMs) to accommodate the high-dimensional
(p\textgreater\textgreater n) problem using group penalties and priors.
These models rarely smooth the nonlinear effect properly because of the
sparse regularization. It leads to excess shrinkage on coefficients and
inaccurate prediction. Besides, most of these GAMs consider an
``all-in-all-out'' approach for function selection, rendering them
difficult to answer if nonlinear effects are necessary. While some
Bayesian models can address the shortcomings, using Markov chain Monte
Carlo (MCMC) algorithms for model fitting creates a new challenge,
scalability. Hence, we propose Bayesian hierarchical generalized
additive models as a solution. For smooth and sparse solutions, the
proposed models employ bi-part spike-and-slab spline priors, including
mixture double exponential and mixture normal priors. Two
optimization-based algorithms, EM-Coordinate Descent and EM-Iterative
Weighted Least Squares, are developed for expedited computation.
Simulation studies and metabolomics data analyses demonstrate improved
performance against state-of-art models, mgcv, COSSO and sparse Bayesian
GAM. The software implementation of the proposed models is freely
available via an R package BHAM.}

@@

\pgfplotsset{compat=1.18}
\usetikzlibrary{shapes.geometric, arrows, positioning, calc, matrix, backgrounds, fit}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\tp}{*}
\newcommand{\pr}{\text{Pr}}

@@

Nonparametric regression models are appropriate alternatives to the
``black-box'' models thanks to their balance between model flexibility
and interpretability. Among those, generalized additive models (GAMs),
proposed in the seminal work of Hastie and Tibshirani
\citep{Hastie1987}, grew to be one of the most popular modeling tools.
In a GAM, the response variable, \(Y\), which is assumed to follow some
exponential family distribution \(EF(\mu, \phi)\) with mean \(\mu\) and
dispersion \(\phi\), can be expressed as the summation of smoothing
functions, \(B_j, j = 1, \dots, p\), of a given \(p\)-dimensional vector
of covariates \(\boldsymbol{x}\), written as \[
 E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)),
\] where \(g^{-1}\) is the inverse of a monotonic link function. The
smoothing functions can take many forms and are estimated using a
pseudo-weighted version of the backfitting algorithm
\citep{Breiman1985}. Nevertheless, the classical GAMs cannot fulfill the
increasing analytic demands for high-dimensional data analysis in
biomedical studies.

@@

To address these challenges, we propose a novel Bayesian hierarchical
generalized additive model (BHAM) for high dimensional data analysis.
Specifically, we incorporate smoothing penalties in the model via
re-parameterization of the smoothing function to avoid overly shrinking
basis function coefficients. Smoothing penalties are commonly
implemented in the classical GAMs through smoothing regression splines.
They are quadratic norms of the coefficients and allow locally adaptive
penalties on each smoothing function. A smoothing penalty conditioning
on a smoothing parameter \(\lambda_j\) is a function of the integration
of the second derivative of the spline function, expressed
mathematically as \begin{equation}\label{eq:smoothpen}
  \text{pen}\left[f_j(x)\right] = \lambda_j \int f^{\prime\prime}_j(x)dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j ,
\end{equation} where \(\boldsymbol{S}_j\) is a known smoothing penalty
matrix, and \(\boldsymbol{\beta}_j\) are the basis function
coefficients. Smoothing penalties were also previously used in the
spike-and-slab GAM \citep{Scheipl2012} and the sparsity-smoothness
penalty \citep{Meier2009}. Moreover, incorporating the smoothing penalty
allows the separation of the linear space of smoothing functions, also
know as the null space, from the nonlinear space. We then impose a new
bi-part spike-and-slab spline prior on the smoothing functions for
bi-level selection such that the linear and nonlinear spaces of
smoothing functions can be selected separately. The prior setup
encourages a flexible solution, rending one of three possibilities: no
effect, only linear effect, or nonlinear effect of a predictor. In
addition, two fast computing optimization-based algorithms are developed
for high and ultrahigh-dimensional\citep{Fan2009} settings respectively.

The proposed framework, BHAM, differs from previous spike-and-slab based
GAMs, i.e.~the spike-and-slab GAM \citep{Scheipl2012} and the SB-GAM
\citep{Bai2021} in many ways. First of all, the proposed spike-and-slab
spline prior is a spike-and-slab lasso type prior using independent
mixture double exponential distribution, compared to spike-and-slab GAM.
Spike-and-slab lasso priors provide computational convenience during
model fitting by using optimization algorithms instead of computational
intensive sampling algorithms. They make fitting high- and
ultrahigh-dimensional models more feasible without sacrificing
performance in prediction and variable selection. Secondly, the
optimization algorithms introduced in this article include EM-Coordinate
Descent (EM-CD) algorithm and EM-Iterative weighted least square
(EM-IWLS) algorithm for different utilities. The applications of the two
aforementioned algorithm are fruitful in Bayesian hierarchical models in
high-dimensional data analysis.
\citep{Yi2012, Rockova2014a, Rockova2018} Even though SB-GAM with a
group spike-and-slab lasso prior also uses EM-CD for model fitting, the
proposed EM-IWLS provides uncertainty measures that could be further
used for confidence bound calculation and hypothesis testing. EM-IWLS
complements the lack of inference potential from EM-CD in a median and
high dimensional setting. Furthermore, the proposed model addresses the
incapability of bi-level selection in SB-GAM. Lastly, the R
implementation of the Bayesian hierarchical additive model is available
publicly via BHAM at \url{https://github.com/boyiguo1/BHAM}, make
transnational science more accessible.

@@

Following the GLM notation introduced in Section \ref{sec:intro}, we
have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation} and the data distribution is \begin{equation}
f(y|X\beta, \phi) = \prod\limits^n_{i=1}f(y_i|X_i,\phi),\nonumber
\end{equation} with smoothing functions \(B_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The basis function matrix, i.e.~the design
matrix derived from the smoothing function \(B_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function. For
example, a cubic spline function with \(k\) knots produces a \(k+3\)
elements column vector. Its corresponding smoothing penalty matrix is
denoted as \(\boldsymbol{S}_j\). \(\boldsymbol{\beta}_j\) denotes the
basis function coefficients for the \(j\)th variable, and the matrix
transposing operation is denoted with a superscript \(^T\). To note, the
proposed model can include parametric forms of variables in the model,
treating the parametric function as a special case of the smoothing
function, e.g.~\(B_j(x_j) = x_j\) with the smoothing penalty matrix
defined as \(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

To encourage proper smoothing of the functions, we adopt the idea of
smoothing penalties from smoothing spline models. The basic idea is to
set up a smoothing penalty, described in Equation (\ref{eq:smoothpen}),
in the prior density function, or least squared and likelihood function
in the penalized framework. However, the direct integration of smoothing
penalties into sparsity penalties is not obvious. Marra and Wood
\citep{Marra2011} proposed a re-parameterization procedure to
accommodate the smoothing penalty implicitly. Given the smoothing
penalty matrix is most often symmetric and positive semi-definite for
univariate smoothing functions, we apply eigendecomposition on the
penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\) ,
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
\(0\) elements on the diagonal, associated with the linear space of the
smoothing function. We further write the orthonormal matrix
\(\boldsymbol{U} \equiv \begin{bmatrix}\boldsymbol{U}^0 : \boldsymbol{U}^{*}\end{bmatrix}\)
containing the eigenvectors as columns in the corresponding order. That
is \(\boldsymbol{U}^0\) contains the eigenvectors with zero eigenvalues
for the linear space and \(\boldsymbol{U}^{*}\) contains the
eigenvectors (as columns) for the non-zero eigenvalues, i.e.~the
non-linear space. We multiply the basis function matrix
\(\boldsymbol{X}\) with the orthonormal matrix \(\boldsymbol{U}\) for
the new design matrix
\(\boldsymbol{X}^\ast = \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{X}^0 : \boldsymbol{X}^{*} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{*}\) by the
non-zero eigenvalues of \(\boldsymbol{D}\) such that the new basis
function matrix \(\boldsymbol{X}^\ast\) can receive uniform penalty on
each of its dimensions. With slight abuse of the notation, we denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} \boldsymbol{X}_j^0 : \boldsymbol{X}_j^{*} \end{bmatrix}\)
as the basis function matrix for the \(j\)th variable after the
re-parameterization. A spline function can be expressed in the matrix
form \[
B_j(x_j) = B_j^0(x_j) + B_j^*(x_j) = \boldsymbol{\beta_j^0}^T \boldsymbol{X}_j^0 + \boldsymbol{\beta_j^*}^T \boldsymbol{X}_j^*,
\] and the generalized additive model in Equation (\ref{eq:gam}) be
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) =  g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[a + \sum\limits^p_{j=1} (\boldsymbol{\beta_j^0}^T \boldsymbol{X}_j^0 + {\boldsymbol{\beta}_j^*}^T \boldsymbol{X}_j^*)\right].
\end{equation}

@@

The major criticism of early spike-and-slab priors receive is being
computationally prohibitive. \citep{Bai2021Review} Since then, many
studies focus on alleviating the computational burden that sampling
algorithms bear, which include EMVS \citep{Rockova2014a} and
Spike-and-slab Lasso \citep{Rockova2018b, Rockova2018}. Particularly,
the development of Spike-and-slab Lasso substantially improves the
scalability of SS models, setting up the theoretical foundation for
generalized models in -omics data analysis
\citep{Tang2017a, Tang2017, Tang2018, Tang2019}. The spike-and-slab
lasso prior is composed of two double exponential distributions with
mean 0 and different dispersion parameters, \(0 < s_0 < s_1\),
mathematically, \begin{equation} 
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.\nonumber
\end{equation} Given that both double exponential distributions have a
mean of 0 and the latent indicator \(\gamma\) can only take the value of
0 or 1, the mixture double exponential distribution can be formulated as
a double exponential density, \begin{equation} \label{eq:ssl}
\beta | \gamma \sim DE(0, (1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1), 0 < s_0 < s_1,
\end{equation} with the scale parameter
\(S = (1-\gamma)s_0 + \gamma s_1\). The SSL also mitigates the problem
of EMVS where the weak signals are not shrink to zero, and hence is
preferred in high-dimensional data analysis. We notice that Bai
\citep{Bai2021} is the first to apply spike-and-slab lasso prior in the
GAM framework, where the densities of the spike and slab components take
the group lasso density \citep{Xu2015}. Bai \citep{Bai2021} takes an
``all-in-all-out'' strategy for function selection.

@@

We introduce a novel prior for GAMs, particularly for high-dimensional
nonlinear modeling with bi-level selection. The proposed prior extends
from the spike-and-slab lasso prior described in Equation
(\ref{eq:ssl}). Given the re-parameterized deign matrix
\(\boldsymbol{X}_j = \begin{bmatrix} \boldsymbol{X}_j^0 : \boldsymbol{X}_j^*\end{bmatrix}\)
for the \(j\)th variable, we impose a two-part SSL prior to the
coefficients \(\boldsymbol{\beta}_j\). Specifically, we impose
independent group priors on the linear space coefficients and on the
nonlinear space coefficients respectively,

\begin{align}\label{eq:bham_ssl}
  \beta^0_{j} |\gamma^0_{j},s_0,s_1 &\sim DE(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1) \nonumber \\
  \beta^*_{jk} | \gamma^*_{j},s_0,s_1 &\sim DE(0,(1-\gamma^*_{j}) s_0 + \gamma^*_{j} s_1), 
\end{align}

where \(\gamma^0_{j}\in\{0,1\}\) and \(\gamma^*_{j}\in \{0,1\}^{K_j}\)
are latent indicator variable and vectors, indicating if the model
includes the linear effect and the nonlinear effect of the \(j\)th
variable respectively. \(s_0\) and \(s_1\) are scale parameters,
assuming \(0 < s_0 < s_1\) and given. These scale parameters \(s_0\) and
\(s_1\) can be treated as tuning parameters and optimized via
cross-validation. A discussion of how to choose the scale parameters
comes in Section \ref{sec:tune}. To note, this prior differs from
previous group spike-and-slab lasso priors \citep{Tang2018, Tang2019},
where the \(\boldsymbol{\beta}^0_j\) and \(\boldsymbol{\beta}^*_j\)
shares the same indicator variables \(\gamma_j^0\), \(\gamma_j^*\)
respectively. It is possible to add a more restrictive assumption on the
priors, assuming that one indicator variable decides the inclusion of
both the linear effect and nonlinear effect,
i.e.~\(\gamma_j = \gamma^0_j = \gamma^*_j\). This converges to the
SB-GAM \citep{Bai2021}. Conversely, it is also possible to relax the
assumption such that each coefficient \(\beta\) to have its own latent
indicator \(\gamma\), but at the cost of complicating the bi-level
function selection. This reduces the proposed prior to the classic
spike-and-slab lasso prior.

@@

The rest of the hierarchical prior follows the traditional
spike-and-slab lasso prior: we set up hyper-priors of
\(\boldsymbol{\gamma}\) to allow local adaption of the shrinkage using a
Bernoulli distribution (binomial distribution of one trial). The two
indicators of the \(j\)th predictor, \(\gamma^{0}_j\) and
\(\gamma^*_j\), shares the same probability parameter \(\theta_j\), \[
\begin{aligned}
\gamma_{j}^{0} | \theta_j &\sim Bin(\gamma^{0}_{j}|1, \theta_j)\\
\gamma_{j}^*| \theta_j &\sim Bin(\gamma^*_{j}|1, \theta_j).
\end{aligned}
\] This is to leverage the fact that the probability of selecting the
bases of a smoothing function should be similar, while allowing
different penalty on the linear space and non-linear space of the
smoothing function. The hyper prior of \(\gamma_{j}^{0}\) decides the
sparsity of the model at the function selection level, while that of
\(\gamma_{j}^*\) decides the smoothness of the spline function at basis
function level. Meanwhile, we specify that \(\gamma_{j}^0\) and
\(\gamma_{j}^*\) are independently distributed for analytic simplicity.
We further specify the parameter \(\theta_j\) follows a beta
distribution with given shape parameters \(a\) and \(b\), \[
\theta_j \sim Beta(a, b).
\] The beta distribution is a conjugate prior for the binomial
distribution and hence provides some computation convenience. For
simplicity, we focus on a special case of beta distribution, uniform
(0,1). When the variable have large effects in any of the bases, the
parameter \(\theta_j\) will be estimated large, which in turn encourages
the model to include the rest of bases, achieving the local adaption
among spline bases. Hereafter, we refer Bayesian hierarchical
generalized additive models with the spike-and-slab lasso prior as the
BHAM-SSL, and visually presented in Figure \ref{fig:SSprior}.

@@

\matrix (linearPrior) [matrix of nodes, column sep = 0mm, row sep = 0.7cm] {
  \node (linearGamma) [outCome] { $\gamma_j^0 \sim Bin(1, \theta_j) $ };\\
  \node (linearBeta) [outCome] { $\beta_j^0 \sim DE(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1)$};\\
};
\matrix (penPrior) [right = 2cm of linearPrior, matrix of nodes, column sep = 0mm, row sep = 0.7cm] {
  \node (penGamma) [outCome] { $\gamma_{j}^*\sim Bin(1, \theta_j)$ };\\
  \node (penBeta) [outCome] { $\beta_{jk}^*\sim  DE(0,(1-\gamma^*_{j}) s_0 + \gamma^*_{j} s_1)$};\\
};

>>>>

\abstract{There are proposals that extend the classical generalized
additive models (GAMs) to accommodate the high-dimensional (\(p>>n\))
problem using group penalties and priors. These models rarely smooth the
nonlinear effect properly because of the sparse regularization. It leads
to excess shrinkage on coefficients and inaccurate prediction. Besides,
most of these GAMs consider an ``all-in-all-out'' approach for function
selection, rendering them difficult to answer if nonlinear effects are
necessary. While some Bayesian models can address the shortcomings,
using Markov chain Monte Carlo (MCMC) algorithms for model fitting
creates a new challenge, scalability. Hence, we propose Bayesian
hierarchical generalized additive models as a solution. For smooth and
sparse solutions, the proposed models employ bi-part spike-and-slab
spline priors, including mixture double exponential and mixture normal
priors. Two optimization-based algorithms, EM-Coordinate Descent and
EM-Iterative Weighted Least Squares, are developed for expedited
computation. Simulation studies and metabolomics data analyses
demonstrate improved performance against state-of-art models, mgcv,
COSSO and sparse Bayesian GAM. The software implementation of the
proposed models is freely available via an R package BHAM.}

@@

\pgfplotsset{compat=1.18}
\usetikzlibrary{shapes.geometric, arrows, positioning, calc, matrix, backgrounds, fit}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\tp}{*}
\newcommand{\pr}{\text{Pr}}
\newcommand{\repa}{\text{repa}}
\newcommand{\simiid}{\overset{\text{iid}}{\sim}}

@@

Nonparametric regression models are appropriate alternatives to the
``black-box'' models thanks to their balance between model flexibility
and interpretability. Among those, generalized additive models (GAMs),
proposed in the seminal work of Hastie and Tibshirani
\citep{Hastie1987}, grew to be one of the most popular modeling tools.
In a GAM, the response variable, \(Y\), which is assumed to follow some
exponential family distribution \(EF(\mu, \phi)\) with mean \(\mu\) and
dispersion \(\phi\), can be modeled with the summation of smoothing
functions, \(B_j(\cdot), j = 1, \dots, p\), of a given \(p\)-dimensional
vector of covariates \(\boldsymbol{x}\), written as \[
 E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)),
\] where \(g^{-1}(\cdot)\) is the inverse of a monotonic link function.
The smoothing functions can take many forms and are estimated using a
pseudo-weighted version of the backfitting algorithm
\citep{Breiman1985}. Nevertheless, the classical GAMs cannot fulfill the
increasing analytic demands for high-dimensional data analysis in
biomedical studies.

@@

To address these challenges, we propose a novel Bayesian hierarchical
generalized additive model (BHAM) for high dimensional data analysis.
Specifically, we incorporate smoothing penalties in the model via
re-parameterization of the smoothing function to avoid overly shrinking
basis function coefficients. Smoothing penalties are commonly
implemented in the classical GAMs through smoothing regression splines.
They are quadratic norms of the coefficients and allow locally adaptive
penalties on each smoothing function. A smoothing penalty conditioning
on a smoothing parameter \(\lambda_j\) is a function of the integration
of the second derivative of the spline function, expressed
mathematically as \begin{equation}\label{eq:smoothpen}
  \text{pen}\left[B_j(x)\right] = \lambda_j \int B^{\prime\prime}_j(x)dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j ,
\end{equation} where \(\boldsymbol{S}_j\) is a known smoothing penalty
matrix, and \(\boldsymbol{\beta}_j\) are the basis function
coefficients. Smoothing penalties were also previously used in the
spike-and-slab GAM \citep{Scheipl2012} and the sparsity-smoothness
penalty \citep{Meier2009}. Moreover, incorporating the smoothing penalty
allows the separation of the linear space of smoothing functions from
the nonlinear space. We then impose a new bi-part spike-and-slab spline
prior on the smoothing functions for bi-level selection such that the
linear and nonlinear spaces of smoothing functions can be selected
separately. The prior setup encourages a flexible solution, rending one
of three possibilities: no effect, only linear effect, or nonlinear
effect of a predictor. In addition, two scalable optimization-based
algorithms are developed for high and
ultrahigh-dimensional\citep{Fan2009} settings respectively.

The proposed framework, BHAM, differs from previous spike-and-slab based
GAMs, i.e.~the spike-and-slab GAM \citep{Scheipl2012} and the SB-GAM
\citep{Bai2021} in many ways. First of all, the proposed spike-and-slab
spline prior is a spike-and-slab lasso type prior using independent
mixture double exponential distribution, compared to spike-and-slab GAM.
Spike-and-slab lasso priors provide computational convenience during
model fitting by using optimization algorithms instead of computational
intensive sampling algorithms. They make fitting high- and
ultrahigh-dimensional models more feasible without sacrificing
performance in prediction and variable selection. Secondly, the
optimization algorithms introduced in this article include EM-Coordinate
Descent (EM-CD) algorithm and EM-Iterative weighted least square
(EM-IWLS) algorithm for different utilities. The applications of the two
aforementioned algorithms are fruitful in Bayesian hierarchical models
for high-dimensional data. \citep{Yi2012, Rockova2014a, Rockova2018}
Even though SB-GAM with a group spike-and-slab lasso prior also uses
EM-CD for model fitting, the proposed EM-IWLS provides uncertainty
measures that could be further used for confidence bound calculation and
hypothesis testing. EM-IWLS complements the lack of inference potential
from EM-CD in a median and high dimensional setting. Furthermore, the
proposed model addresses the incapability of bi-level selection in
SB-GAM. Lastly, the R implementation of the Bayesian hierarchical
additive model is available publicly via BHAM at
\url{https://github.com/boyiguo1/BHAM}, making transnational science
more accessible.

@@

Following the GLM notation introduced in Section \ref{sec:intro}, we
have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation} and the data distribution is \begin{equation}
f(\boldsymbol{y}| \boldsymbol{\beta}, \phi) = \prod\limits^n_{i=1}f(y_i|\boldsymbol{\beta}, \phi),\nonumber
\end{equation} with smoothing functions \(B_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The basis function matrix, i.e.~the design
matrix derived from the smoothing function \(B_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function. For
example, a cubic spline function with \(k\) knots produces a \(k+3\)
elements column vector. Its corresponding smoothing penalty matrix is
denoted as \(\boldsymbol{S}_j\). \(\boldsymbol{\beta}_j\) denotes the
basis function coefficients for the \(j\)th variable such that
\(B_j(x_j) = \boldsymbol{\beta}_j^T \boldsymbol{X}_j\). With slight
abuse of notation, we denote vectors and matrices in bold fonts
\(\boldsymbol{\beta}, \boldsymbol{X}\) with conformable dimensions,
where scalar and random variables are denoted in with unbold fonts
\(\beta, X\). The matrix transposing operation is denoted with a
superscript \(^T\). To note, the proposed model can include parametric
forms of variables in the model, treating the parametric function as a
special case of the smoothing function, e.g.~\(B_j(x_j) = x_j\) with the
smoothing penalty matrix defined as
\(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

To encourage proper smoothing of the functions, we adopt the idea of
smoothing penalties from smoothing spline models. The basic idea is to
set up a smoothing penalty, described in Equation (\ref{eq:smoothpen}),
in the prior density function. However, the direct integration of
smoothing penalty with sparsity penalty is not obvious. Marra and Wood
\citep{Marra2011} proposed a re-parameterization procedure to
accommodate the smoothing penalty implicitly. Given the smoothing
penalty matrix \(\boldsymbol{S}_j\) is symmetric and positive
semi-definite for univariate smoothing functions, we apply
eigendecomposition on the penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\) ,
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
elements of zeros on the diagonal, where the zeros are associated with
the linear space of the smoothing function. For the most popular
smoothing function, cubic splines, the dimension of the linear space is
one. Hereafter, we focus on discussing a uni-dimensional linear space
for simplicity; however, it generalizes easily to the cases where the
linear space is multidimensional. We further write the orthonormal
matrix
\(\boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{U}^0 : \boldsymbol{U}^{*}\end{bmatrix}\)
containing the eigenvectors as columns in the corresponding order to
\(\boldsymbol{D}\). That is, \(\boldsymbol{U}\) contains the
eigenvectors \(U^0\) with zero eigenvalues for the linear space and
\(\boldsymbol{U}^{*}\) contains the eigenvectors (as columns) for the
non-zero eigenvalues, i.e.~the non-linear space. We multiply the basis
function matrix \(\boldsymbol{X}\) with the orthonormal matrix
\(\boldsymbol{U}\) for the new design matrix
\({\boldsymbol{X}}^\text{repa}= \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} X^0 : \boldsymbol{X}^{*} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{*}\) by the
non-zero eigenvalues of \(\boldsymbol{D}\) such that the new basis
function matrix \(\boldsymbol{X}^\ast\) can receive uniform penalty on
each of its dimensions. With slight abuse of the notation, we drop the
superscript \(^\text{repa}\) and denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} X_j^0 : \boldsymbol{X}_j^{*} \end{bmatrix}\)
as the basis function matrix for the \(j\)th variable after the
re-parameterization. A spline function can be expressed in the matrix
form \[
B_j(x_j) = B_j^0(x_j) + B_j^*(x_j) = \beta_j X^0_j + \boldsymbol{\beta_j^*}^T \boldsymbol{X}_j^*,
\] and the generalized additive model in Equation (\ref{eq:gam}) be
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1} B_j(x_j)) = g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[a + \sum\limits^p_{j=1} (\beta_j X^0_j + {\boldsymbol{\beta}_j^*}^T \boldsymbol{X}_j^*)\right],
\end{equation} where the coefficients
\(\boldsymbol{\beta}_j \equiv \begin{bmatrix} \beta_j : \boldsymbol{\beta}^*_j \end{bmatrix}\)
is an augmentation of the coefficient scalar \(\beta_j\) of linear space
and the coefficient vector \(\boldsymbol{\beta}^*_j\) of non-linear
space.

@@

The major criticism of early spike-and-slab models is being
computationally prohibitive. \citep{Bai2021Review} Since then, many
studies focus on alleviating the computational burden that sampling
algorithms bear, which include EMVS \citep{Rockova2014a} and
spike-and-slab Lasso \citep{Rockova2018b, Rockova2018}. Particularly,
the development of spike-and-slab Lasso substantially improves the
scalability of SS models, setting up the theoretical foundation for
generalized models in -omics data analysis
\citep{Tang2017a, Tang2017, Tang2018, Tang2019}. The spike-and-slab
Lasso (SSL) prior is composed of two double exponential distributions
with mean 0 and different dispersion parameters, \(0 < s_0 < s_1\),
mathematically, \begin{equation} 
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.\nonumber
\end{equation} Given that both double exponential distributions have a
mean of 0 and the latent indicator \(\gamma\) can only take the value of
0 or 1, the mixture double exponential distribution can be formulated as
one single double exponential density, \begin{equation} \label{eq:ssl}
\beta | \gamma \sim DE(0, S), 0 < s_0 < s_1,
\end{equation} with the scale parameter
\(S = (1-\gamma)s_0 + \gamma s_1\). The SSL also mitigates the problem
of EMVS where the weak signals are not shrink to zero, and hence is
preferred in high-dimensional data analysis. We notice that Bai
\citep{Bai2021} is the first to apply spike-and-slab lasso prior in the
GAM framework, where the densities of the spike and slab components take
the group lasso density \citep{Xu2015}. Bai \citep{Bai2021} takes an
``all-in-all-out'' strategy for function selection.

@@

We introduce a novel prior for GAMs, particularly for high-dimensional
nonlinear modeling with bi-level selection. The proposed prior extends
from the spike-and-slab lasso prior described in Equation
(\ref{eq:ssl}). Given the re-parameterized deign matrix
\(\boldsymbol{X}_j = \begin{bmatrix} X^0_j : \boldsymbol{X}_j^*\end{bmatrix}\)
for the \(j\)th variable, we impose a two-part SSL prior to the
coefficients
\(\boldsymbol{\beta}_j = \begin{bmatrix} \beta_j : \boldsymbol{\beta}_j^*\end{bmatrix}\).
Specifically, we impose independent group priors on the linear space
coefficients and on the nonlinear space coefficients respectively,
\begin{align}\label{eq:bham_ssl}
  \beta_{j} |\gamma_{j},s_0,s_1 &\sim DE(0,(1-\gamma_{j}) s_0 + \gamma_{j} s_1) \nonumber \\
  \beta^*_{jk} | \gamma^*_{j},s_0,s_1 &\overset{\text{iid}}{\sim}DE(0,(1-\gamma^*_{j}) s_0 + \gamma^*_{j} s_1), k=1,\dots, K_j
\end{align} where \(\gamma_{j}\in\{0,1\}\) and
\(\gamma^*_{j}\in \{0,1\}\) are two latent indicator variables,
indicating if the model includes the linear effect and the nonlinear
effect of the \(j\)th variable respectively. \(s_0\) and \(s_1\) are
scale parameters, assuming \(0 < s_0 < s_1\) and given. These scale
parameters \(s_0\) and \(s_1\) can be treated as tuning parameters and
optimized via cross-validation. A discussion of how to choose the scale
parameters comes in Section \ref{sec:tune}. To note, this prior differs
from previous group spike-and-slab lasso priors
\citep{Tang2018, Tang2019}, as the \(\beta_j\) and
\(\boldsymbol{\beta}^*_j\) have different indicator variables
\(\gamma_j\), \(\gamma_j^*\) respectively. It is possible to add a more
restrictive assumption on the priors, assuming that one indicator
variable decides the inclusion of both the linear effect and nonlinear
effect, i.e.~\(\gamma_j = \gamma^*_j\). This converges to the SB-GAM
\citep{Bai2021}. Conversely, it is also possible to relax the assumption
such that each coefficient \(\beta_{jk} \in \boldsymbol{\beta}^*_j\) has
its own latent indicator \(\gamma_{jk}\), but at the cost of
complicating the bi-level function selection. This reduces the proposed
prior to the classic spike-and-slab lasso prior.

@@

The rest of the hierarchical prior follows the traditional
spike-and-slab lasso prior: we set up hyper-priors on
\(\gamma_j, \gamma^*_j\) to allow local adaption of the shrinkage using
a Bernoulli distribution, written as binomial distribution of one trial.
The two indicators of the \(j\)th predictor, \(\gamma_j\) and
\(\gamma^*_j\), shares the same probability parameter \(\theta_j\), \[
\begin{aligned}
&\gamma_{j} | \theta_j \sim Bin(1, \theta_j) & & 
&\gamma_{j}^*| \theta_j \sim Bin(1, \theta_j).
\end{aligned}
\] This is to leverage the fact that the probability of selecting the
bases of a smoothing function should be similar, while allowing
different penalty on the linear space and non-linear space of the
smoothing function. The hyper prior of \(\gamma_{j}\) decides the
sparsity of the model at the function selection level, while that of
\(\gamma_{j}^*\) decides the smoothness of the spline function at basis
function level. Meanwhile, we specify that \(\gamma_{j}\) and
\(\gamma_{j}^*\) are independently distributed for analytic simplicity.
We further specify the parameter \(\theta_j\) follows a beta
distribution with given shape parameters \(a\) and \(b\), \[
\theta_j \sim Beta(a, b).
\] The beta distribution is a conjugate prior for the binomial
distribution and hence provides some computation convenience. For
simplicity, we focus on a special case of beta distribution, uniform
(0,1), i.e.~\(a = 1, b = 1\). When the variable have large effects in
any of the bases, the parameter \(\theta_j\) will be estimated large,
which in turn encourages the model to include the rest of bases,
achieving the local adaption among spline bases. Hereafter, we refer
Bayesian hierarchical generalized additive models with the
spike-and-slab lasso prior as the BHAM-SSL, and visually presented in
Figure \ref{fig:SSprior}.

@@

\matrix (linearPrior) [matrix of nodes, column sep = 0mm, row sep = 0.7cm] {
  \node (linearGamma) [outCome] { $\gamma_j \sim Bin(1, \theta_j) $ };\\
  \node (linearBeta) [outCome] { $\beta_j \sim DE(0,(1-\gamma_{j}) s_0 + \gamma_{j} s_1)$};\\
};
\matrix (penPrior) [right = 2cm of linearPrior, matrix of nodes, column sep = 0mm, row sep = 0.7cm] {
  \node (penGamma) [outCome] { $\gamma_{j}^*\sim Bin(1, \theta_j)$ };\\
  \node (penBeta) [outCome] { $\beta_{jk}^*\sim  DE(0,(1-\gamma^*_{j}) s_0 + \gamma^*_{j} s_1)$};\\
};

\node (s) [staticCompo]  at ($(linearBeta)!0.5!(penBeta)$)  {($s_0, s_1$)};
\node (Beta) [staticCompo, below = 1cm of s] {$\boldsymbol{\beta }= (\beta_1, \boldsymbol{\beta}^*_1, \dots,\beta_j, \boldsymbol{\beta}^*_j , \dots,\beta_p, \boldsymbol{\beta}^*_p) $};
\node (Theta)[outCome, above = 2cm of s] {$\theta_{j} \sim Beta(a, b)$};
\node (ab)[staticCompo, above = 0.5cm of Theta] {$(a, b)$};
\node (Y) [outCome, below = 1cm of Beta] {$y_i \sim Expo. Fam. (g^{-1}(\boldsymbol{\beta}^T \boldsymbol{X}_i), \phi)$};
